Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2015-06-04 tex/camera_ready/paper.tex
minor edits

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
 each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$
and each variable $x:\tau$ of $\svars{\Sigma}$ as an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(that is, for every $I \in \Mo$, every $\Sigma$-interpretation that differs
from $I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{longv};
otherwise, it is \emph{$T$-unsatisfiable}\end{longv}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
%ct changed the following because it is not adequate for our purposes
%
%Given a signature $\Sigma$,
%the \emph{theory of equality with uninterpreted functions} \euf
%consists of the set of all $\Sigma$-interpretations.
%Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%%\ct{This is both ambiguous and insufficient.
%%Since this is SMT we should fix some background theory $T$,
%%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%%}
If $T_1 = (\Sigma_1, \Mo_1)$ is a theory and $\Sigma_2$ is a signature 
with $\sfuns{\Sigma_1} \cap \sfuns{\Sigma_2} = \emptyset$,
the \emph{extension of $T_1$ to $\Sigma_2$} is the theory $T = (\Sigma, \Mo)$ where 
$\sfuns{\Sigma} = \sfuns{\Sigma_1} \cup \sfuns{\Sigma_2}$,
$\stypes{\Sigma} = \stypes{\Sigma_1} \cup \stypes{\Sigma_2}$,
and $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_1$-reduct is a model of $T_1$.
We refer to the symbols of $\Sigma_2$ that are not is $\Sigma_1$ as \emph{uninterpreted}.
For the rest of the paper \emph{we fix a theory $T = (\Sigma, \Mo)$ 
with uninterpreted symbols} constructed as above.

We will consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
uninterpreted. Their
semantics is the same as for standard quantified formulas
$\forall \vec x.\; \varphi$.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1 : \tau_1$, $\ldots,$ $x_n : \tau_n$
and $\varphi$ is a quantifier-free formula 
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
%\ct{Why can't $\varphi$ be existential instead of just quantifier free?} 
%I don't think it makes much sense to have any quantifiers within function bodies, 
%since they aren't operational, that's why I put quantifier-free. --ajr
%%\jb{Actually, in Isabelle we can have them and we use them sometimes. It certainly
%% makes sense. I can give you real examples if you like.}
We write $\exists
\vec x.\; \varphi$ as an abbreviation for $\neg \forall \vec x.\; \neg \varphi$.

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples to formulas involving recursive
function definitions.
For convenience, it
has been implemented in the development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\begin{longv}%
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{longv}%
\begin{shortv}%
\cvc adds
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$. \end{shortv}%
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the translation $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs a partial model of~$\con{s}$, where only
the values of $\con{s}$ for $x = 0,\ldots,14$ are given.

>>>>

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
 each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ as an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(i.e., for every $I \in \Mo$, every $\Sigma$-interpretation that differs
from $I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{longv};
otherwise, it is \emph{$T$-unsatisfiable}\end{longv}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
%ct changed the following because it is not adequate for our purposes
%
%Given a signature $\Sigma$,
%the \emph{theory of equality with uninterpreted functions} \euf
%consists of the set of all $\Sigma$-interpretations.
%Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%%\ct{This is both ambiguous and insufficient.
%%Since this is SMT we should fix some background theory $T$,
%%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%%}
If $T_1 = (\Sigma_1, \Mo_1)$ is a theory and $\Sigma_2$ is a signature 
with $\sfuns{\Sigma_1} \cap \sfuns{\Sigma_2} = \emptyset$,
the \emph{extension of $T_1$ to $\Sigma_2$} is the theory $T = (\Sigma, \Mo)$ where 
$\sfuns{\Sigma} = \sfuns{\Sigma_1} \cup \sfuns{\Sigma_2}$,
$\stypes{\Sigma} = \stypes{\Sigma_1} \cup \stypes{\Sigma_2}$,
and $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_1$-reduct is a model of $T_1$.
We refer to the symbols of $\Sigma_2$ that are not in $\Sigma_1$ as \emph{uninterpreted}.
For the rest of the paper, \relax{we fix a theory $T = (\Sigma, \Mo)$ 
with uninterpreted symbols} constructed as above.

We consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
uninterpreted. Their
semantics is the same as for standard quantified formulas
$\forall \vec x.\; \varphi$.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1 : \tau_1$, $\ldots,$ $x_n : \tau_n$
and $\varphi$ is a quantifier-free formula 
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
%\ct{Why can't $\varphi$ be existential instead of just quantifier free?} 
%I don't think it makes much sense to have any quantifiers within function bodies, 
%since they aren't operational, that's why I put quantifier-free. --ajr
%%\jb{Actually, in Isabelle we can have them and we use them sometimes. It certainly
%% makes sense. I can give you real examples if you like.}
We write $\exists
\vec x.\; \varphi$ as an abbreviation for $\neg \forall \vec x.\; \neg \varphi$.

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples to formulas involving recursive
function definitions.
For convenience, it
has been implemented in the development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\begin{longv}%
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{longv}%
\begin{shortv}%
\cvc adds
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$. \end{shortv}%
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the translation $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs a partial model of~$\con{s}$ where only
the values of $\con{s}(x)$ for $x = 0,\ldots,14$ are given.
